Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-07 paper.Rnw

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, development, and
neuropsychiatric disorders \citep{Mouiha2011b,Sabuncu2011}.  The result is that
existing hippocampal atlases available to a researcher may not accurately
represent neuroanatomy of a specific population under study.  Additionally, in
the course of a research or clinical study, it may be necessary to make
adjustments to hippocampal definition as a means of hypothesis testing.  For
example, Poppenk \citep{Poppenk2011} found that subdividing the hippocampus
into anterior and posterior regions resulted in a predictive relationship
between volume difference of those regions and recollection memory performance.
Making modifications to an set of atlas library segmentations may require the
same level of expertise and manual effort needed to create the segmentations
from scratch.

@@

The multi-atlas segmentation technique was pioneered by Heckemann et al. with
the MAPER algorithm \citep{Heckemann2006,Heckemann2011}. The basic strategy is as
follows. Each atlas is warped to fit a subject's neuroanatomy using nonlinear
registration techniques \citep{Collins1995,Klein2009}. Atlas labels are then
transformed by this warping to the subject space and a {\it label fusion}
technique, such as voxel-wise voting, is used to merge the label definitions
from each atlas into a definitive segmentation. Typically, an atlas selection
approach is used to improve accuracy by only using labels from atlases that are
most similar to a subject image (cross-correlation of image intensities is a
common similarity metric) \citep{Aljabar2009}.

@@

The insight of generating a template library is not new.  Heckemann et al.
compared "indirect" segmentation -- taking a single atlas and propagating the
labels to intermediate subjects before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is demonstrate the viability of
achieving comparable segmentation accuracy using the basic multi-atlas schema
and using significantly fewer manually created atlases.

>>>>

Compounding each of these problems is the significant neuroanatomical
variability in the hippocampus throughout the course of aging, maturation, and
neuropsychiatric disorders
\citep{Mouiha2011b,Sabuncu2011,Giedd1998,Gogtay2006,Narr2002}.  The result is that
existing hippocampal atlases available to a researcher may not accurately
represent neuroanatomy of a specific population under study.  Additionally, in
the course of a research or clinical study, it may be necessary to make
adjustments to hippocampal definition as a means of hypothesis testing.  For
example, Poppenk \citep{Poppenk2011} found that subdividing the hippocampus
into anterior and posterior regions resulted in a predictive relationship
between volume difference of those regions and recollection memory performance.
Making such modifications to a set of MRI data segmentations requires additional
manual effort.

@@

The multi-atlas segmentation technique was pioneered by Heckemann et al. with
the MAPER algorithm \citep{Heckemann2006,Heckemann2011}. The basic strategy is as
follows. Each atlas image is fit a subject's neuroanatomy using nonlinear
registration techniques \citep{Collins1995,Klein2009}. This resulting nonlinear
transformation is then applied to the atlas labels to bring them into the
subject space.  Finally, a {\it label fusion} technique, such as voxel-wise
voting, is used to merge these labels into a definitive segmentation for the
subject. Typically, an atlas selection approach is used to improve accuracy by
only using labels from atlases that are most similar to a subject image
(cross-correlation of image intensities is a common similarity metric)
\citep{Aljabar2009}.

@@

The insight of generating a template library is not new.  Heckemann et al.
compared ``indirect'' segmentation -- taking a single atlas and propagating the
labels to intermediate subjects before fusing them in a target image space -- to
multi-atlas segmentation and found that the indirect approach performed worse
\citep{Heckemann2006}. In this paper we continue the same line of investigation
but explore the performance when using multiple atlases as well as the effect of
different registration and fusion methods.

The LEAP algorithm \citep{Wolz2010}, described above, is another example of
indirect segmentation previously explored.  LEAP proceeds by iteratively
segmenting unlabelled images most similar to the atlas library images and then
incorporating the labelled images into the atlas library for future iterations.
The novelty explored in our current work is to demonstrate the viability of
achieving comparable segmentation accuracy using the basic multi-atlas schema
and using significantly fewer manually created atlases.
